---
title: "Challenge 2: Spicing things up with ggplot2"
author: "Marietta Nikolskaya"
format: html
editor: source
embed-resources: true
code-tools: true
---

For this week's Challenge, you will have two different options to explore.
I've arranged these options in terms of their "spiciness," or the difficulty of 
completing the task. **You are only required to complete one task, but if you are interested in exploring more than one, feel free!**

This is a great place to let your creativity show! Make sure to indicate what
additional touches you added, and provide any online references you used.

# Directions

For this week's Challenge, you will have two different options to explore.
I've arranged these options in terms of their "spiciness," or the difficulty of 
completing the task. **You are only required to complete one task, but if you are interested in exploring more than one, feel free!**

This is a great place to let your creativity show! Make sure to indicate what
additional touches you added, and provide any online references you used.

::: callout-tip
For this challenge you will be modifying the code you wrote for Lab 2 -- we will
be modifying the box plot of weights by species!
:::

# Data

In the code chunk below, load in the packages and data necessary for this 
analysis. *Hint: You should be able to copy and paste the set-up code from your Lab 2!*
```{r}
library(tidyverse)     
library(dplyr)
library(ggplot2)

#| label: data-packages
surveys <- read_csv("surveys.csv")
glimpse(surveys)

```


# Medium: Exploring Colors

The built-in `ggplot()` color scheme may not include the colors you were looking
for. Don't worry -- there are many other color palettes available to use!

You can change the colors used by `ggplot()` in a few different ways.

### Manual Specification

Add the `scale_color_manual()` or `scale_fill_manual()` functions to your plot
and directly specify the colors you want to use. You can either:

1.  define a vector of colors within the `scale` functions (e.g.
`values = c("blue", "black", "red", "green")`) OR

2.  create a vector of colors using hex numbers and store that vector as a
variable. Then, call that vector in the `scale_color_manual()` function.

### Package Specification

While manual specification may be useful for some contexts, it can be a 
real pain to handpick 5+ colors. This is where color scales built-in to R
packages come in handy! Popular packages for colors include:

-   `RColorBrewer` -- change colors by using `scale_fill_brewer()` or `scale_colour_brewer()`.

-   `viridis` -- change colors by using `scale_colour_viridis_d()` for discrete
data, `scale_colour_viridis_c()` for continuous data.

-   `ggsci` -- change colors by using `scale_color_<PALNAME>()` or
`scale_fill_<PALNAME>()`, where you specify the name of the palette you wish to
use (e.g. `scale_color_aaas()`).

> **For this challenge you are expected to use this information to modify the boxplots you created Lab 2. First, you are to color the boxplots based on the variable `sex`. Next, you are to change the colors used for `sex` using either manual color specification or any of the packages listed here (or others!).**

```{r}
#| label: medium-option-colors
library(RColorBrewer)

ggplot(data = surveys, 
       aes(x = weight, y = species, fill = sex)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(color = "steelblue", alpha = 0.4) +
  labs(x = "Body Weight (grams)", 
       y = "Species", 
       title = "Distribution of Body Weight by Species") +
  scale_fill_brewer(palette = "Set1")

```


# Spicy: Exploring Themes

Themes are another vital component to creating clear plots. During lecture, 
Dr. Theobold mentioned that they like using `theme_bw()` for a built-in theme, 
since it has a white background and easy to see grid lines. However, there are 
SO many more ways you can control the appearance of a plot using the `theme()`
function. 

> **For this challenge you are expected to use the `theme()` function to modify at least two aspects of your plot, with the goal of making your plot clearer to the reader. You need to *explain why* you made the decisions you did.**

Some ideas you might consider:

- changing the size of the font to make your plot titles more readable
- changing the font in your plot to be your favorite font 
(remember it needs to **improve** the plot ;) )
- removing the minor grid lines (since they can be a bit distracting)
- adjust spacing between axis title and labels
- remove axis ticks
- add borders to the plot

```{r}
#| label: spicy-option-theme

ggplot(data = surveys, 
       aes(x = weight, y = species, fill = sex)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(color = "steelblue", alpha = 0.4) +
  labs(x = "Body Weight (grams)", 
       y = "Species", 
       title = "Distribution of Body Weight by Species") +
  scale_fill_manual(values = c("male" = "#1f78b4", "female" = "#e31a1c")) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),   
    panel.grid.minor = element_blank()

  )
```


# Hot: Embedding the Legend in the Plot Title

Some data scientists advocate that we should try to eliminate legends from our
plots to make them more clear. Legends cause a reader's eye to stray from the
plot, which can make it more difficult to understand. There are a few ways to 
do this, but this week we're going to explore how to embed the legend in the 
plot title.

## Embedding Legend in Plot Title

The **ggtext** package allows you to add hex colors and other HTML elements 
(e.g., italics, boldface) to plot titles. The process involves two main steps:

1. Wrap your text in HTML tags within the `labs()` function. 

2. Tell ggplot to render the HTML by adding `plot.title = element_markdown()`
inside the `theme()` function. 

Let's see how this can look! I've added the hex colors for the islands to the
subtitle of my plot:

```{r}
#| echo: true
#| label: html-subtitle
#| eval: false
install.packages("palmerpenguins")
library(palmerpenguins)

ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm, y = species, color = island)
       ) +
  geom_boxplot() +
  scale_color_manual(values = untitled_red) +
  labs(x = "Bill Length (mm)", 
       y = "", 
       title = "Length of Different Penguin Species' Bills", 
       subtitle = "Separated by <span style = 'color:#c8251d;'>Biscoe</span>, <span style = 'color:#a11b26;'>Dream</span> and <span style = 'color:#ef6552;'>Torgenson</span> Islands") +
  theme(
    legend.position = "none", 
    plot.subtitle = element_markdown()
    ) +
  theme_bw()
```

Notice that the subtitle is still specified as a string. Inside the string there
are HTML elements (`<span>`) that declare the colors of the text. For example, 
`<span style = 'color:#c8251d;'>Biscoe</span>` declares that the text "Biscoe"
should be printed with the color `#c8251d`. The beginning of the span (`<span `)
and the end of the span (`</span>`) declare when the coloring starts and ends. 

> **For this challenge you are expected to use this information to modify the boxplots you created in Lab 2. First, you are to color the boxplots based on the variable `sex`. Next, you need to move the legend colors for each level of `sex` into 
the plot title or subtitle. Finally, you are expected to use the `theme()` function to remove the color legend from the plot, since it is no longer needed!**

```{r}

#install.packages("ggtext")
library(ggtext)

#| label: hot-option-removing-plot-title
sex_cols <- c("F" = "#E4572E",  
              "M"   = "#4C78A8") 

ggplot(surveys, aes(x = weight, y = species, fill = sex)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(color = "grey45", alpha = 0.35, width = 0, height = 0.1) +
  labs(
    title = "Distribution of Body Weight by Species",
    subtitle = paste0(
      "Sex: ",
      "<span style='color:#4C78A8;'>Male</span> • ",
      "<span style='color:#E4572E;'>Female</span>"
    ),
    x = "Body Weight (grams)",
    y = "Species"
  ) +
  scale_fill_manual(values = sex_cols) +
  theme_bw() +
  theme(
    legend.position = "none",                         
    plot.subtitle = element_markdown(size = 12),      
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 8)),
    axis.title.y = element_text(margin = margin(r = 8))
  )
```


# Challenge 2 Submission

For Lab 2 you will submit **only** your HTML file. Your HTML file is required to have the following specifications in the YAML options (at the top of your document):

-   have the plots embedded (`embed-resources: true`)
-   include your source code (`code-tools: true`)
-   include all your code and output (`echo: true`)

**If any of the options are not included, your Lab 2 or Challenge 2 assignment will receive an "Incomplete" and you will be required to submit a revision.**

In addition, your document should not have any warnings or messages output in your HTML document. **If your HTML contains warnings or messages, you will receive an "Incomplete" for document formatting and you will be required to submit a revision.**
